\section{Conclusion}\label{sec:conclusion}

We studied decision-focused learning (DFL) for the discrete Berth Allocation Problem (BAP) at the Port of San Antonio, Chile. A predictor estimates each vessel's berth service time, and these estimates feed a mixed-integer linear program (MILP) that schedules a week of berth assignments. Unlike the standard DFL setting, the predicted service time enters the precedence constraints of the MILP rather than a linear objective cost vector. This places the problem in the harder predicted-in-constraints regime, where the Smart Predict-then-Optimize surrogate (SPO+) does not apply. We therefore differentiate through the solver using black-box and perturbed gradients as implemented in PyEPO. By aligning the DFL loss and the regret metric with the solver's unweighted objective, we obtain an exact non-negative-regret guarantee.

Our central hypothesis was cascade asymmetry. Under-predicting a service time propagates delays to downstream vessels, whereas over-predicting only idles a berth. A decision-focused loss should therefore outperform mean squared error (MSE) when schedules are contention-sensitive, and this is what the controlled experiments show. On synthetic instances, DFL reduces schedule regret, and the benefit grows with berth contention: mean regret is about a third lower at moderate contention, while the advantage is negligible or slightly negative at low contention. The gain also depends on an adequate training budget.

On the real 2019--2025 vessel-call data, the picture is different. DFL does not reliably improve on predict-then-optimize. The measured effect lies within run-to-run variance, and the black-box gradient is unstable: validation regret diverges after a few epochs, so early stopping is essential. We report this outcome plainly. The mechanism that DFL exploits on synthetic instances is present in principle in the real data, but the instability of the gradient and the geometry of the observed instances prevent it from translating into a dependable gain.

The contributions of this work are threefold. First, to our knowledge this is the first decision-focused formulation of berth allocation, and it operates in the predicted-in-constraints regime rather than the more common cost-vector regime. Second, the alignment of the loss, the regret metric, and the solver objective yields an exact non-negative-regret guarantee. Third, we provide an honest empirical characterization of when DFL helps, namely under contention, together with the obstacles that remain in moving from synthetic to operational data.

Several directions follow from these findings. The gradient instability observed on real data motivates feasibility-aware and lower-variance estimators \citep{mandi2025feasibility, berthet2020learning}, which may recover on operational instances the benefit seen on synthetic ones. The gap between synthetic and real performance also points to the instance geometry itself: constructing temporally coherent real weekly instances, rather than relying on synthetic instance geometry, would test whether the cascade channel is muted by the structure of the training set or by the estimator. We re-fit the predictors without the suspected post-service feature and found that its removal changes accuracy by only a few percent, so the predictive results do not hinge on it; confirming the feature's measurement time with the port would close the question definitively. Extending the objective with vessel-priority weights would let DFL additionally learn prioritization; the present unweighted objective was chosen deliberately to isolate the cascade channel, and a weighted objective would broaden the decisions the loss can shape. Finally, validation across multiple ports and periods is required before the approach can be recommended for operational use.

Decision-focused learning offers a principled way to train predictors against the decisions they support, and the berth allocation setting exposes both its promise and its current limits. On synthetic instances the cascade-asymmetry hypothesis holds and the benefit scales with contention. On real data the method is not yet a dependable improvement over predict-then-optimize, and closing that gap is the central task for future work.
