\section{Discussion}\label{sec:discussion}

The experiments in Section~\ref{sec:results} give a mixed and, we believe,
instructive picture. Decision-focused learning (DFL) reduces schedule regret in
the synthetic experiments, but only under the conditions the method is designed
to exploit, and the gain does not carry over to the real-data experiment. This
section interprets that pattern through the cascade-asymmetry mechanism of
Section~\ref{sec:method}, explains why the real-data gains fail to materialize,
sets out the threats to validity that qualify every claim above, and states what
the results do and do not imply for practice.

\subsection{When does decision-focused learning help?}\label{sec:disc-when}

The synthetic results are best read through the contention of the berth
schedule. At contention $1.0$, where berth demand is close to capacity, DFL cuts
mean schedule regret by roughly a third relative to predict-then-optimize (PtO).
At contention $0.5$ the two methods are within run-to-run noise, and DFL is
occasionally slightly worse. This dependence is exactly what the
cascade-asymmetry mechanism of Section~\ref{sec:dfl} predicts. When contention is
high, berth assignments are packed back to back, and an under-predicted service
time propagates: the optimizer starts the following vessel too early, the overlap
is realized when the true service time arrives, and the delay cascades down the
berth queue. DFL learns to hedge against this asymmetry by biasing its
predictions upward, trading predictive accuracy for schedules that absorb the
costly error. That hedge only pays off when cascades are frequent. At low
contention vessels rarely conflict, the realized schedule cost is nearly
insensitive to the prediction fed into the optimizer, and there is little
decision signal for the decision-focused loss to exploit; the regret surface is
close to flat in the neighbourhood of the predictions, so a loss defined on that
surface cannot distinguish good hedges from bad ones.

Two secondary observations reinforce this reading. First, the improvement in
decisions coincides with a deterioration in predictive accuracy: the DFL
predictor has worse mean absolute error than its PtO counterpart, consistent with
a model that is no longer trying to match the service time but to shape the
downstream schedule. Better decisions bought with worse forecasts are the
signature of a working decision-focused loss, not an anomaly. Second, the effect
is contingent on adequate training. An under-trained DFL run was worse than PtO,
which indicates that the decision-focused gradient needs enough epochs to move
the predictor away from its warm start before the hedge is expressed. The benefit
of DFL is therefore conditional on both the decision environment and the
optimization budget, not a property the method confers automatically.

\subsection{Why the real-data gains do not materialize}\label{sec:disc-realdata}

In the real-$\tau$ decision experiment DFL does not reliably beat PtO. Across
learning rates the DFL-minus-PtO regret gap lies between roughly $-1\%$ and
$+2\%$, which is smaller than the run-to-run variation in PtO's own regret
(about two units on a scale near thirty). We read this as no reliable
improvement rather than a small one, and several factors plausibly account for
it.

First, the gradient is the limiting factor. Differentiating through the berth
MILP with a black-box estimator produces a high-variance signal, and validation
regret behaves accordingly: it improves for one or two epochs and then diverges,
in one case climbing from about thirty to about seventy. Early stopping is
therefore essential, but it also caps what DFL can achieve, because the best
epoch barely departs from the PtO warm start. This instability is not particular
to our implementation; it is the known difficulty of learning predictors whose
outputs enter the constraints of the downstream program rather than only its
objective \citep{hu2023twostage, mandi2025feasibility}, where small changes in
the predicted parameters can move the feasible region and make the
solution mapping erratic.

Second, the real ``instances'' lack the structure DFL exploits. Because we do not
have coherent weekly berth plans, each instance is assembled by grouping
unrelated real vessels under a synthetic arrival and weight geometry. The
resulting instances reproduce the marginal feature distribution of the port but
not the temporal coherence of an actual berth week, and it is precisely that
coherence that generates the sustained contention on which the cascade mechanism
depends. Weakening the contention structure weakens the only channel through
which DFL improved decisions in the synthetic study.

Third, the effect we are trying to detect is small relative to the noise floor of
the estimate. Validation sets are small, so the regret estimate is itself noisy,
and a genuine gain of one or two percent is difficult to separate from sampling
variation at this scale. Fourth, DFL is warm-started from the PtO solution. In
the absence of a clear decision signal the decision-focused loss has nothing to
pull the predictor toward, so the best it can do is match the baseline, and the
divergent-gradient episodes mean it can also degrade it. Taken together, these
observations are consistent with the broader DFL literature's caution about
predicted parameters that appear in constraints, and they temper any claim that
decision-focused training is a general improvement for berth scheduling on
operational data.

\subsection{Threats to validity}\label{sec:disc-threats}

We record the main threats explicitly, in decreasing order of consequence for the
conclusions.

\emph{Suspected target leakage.} As discussed in Section~\ref{sec:datalimits},
the draft-difference feature depends on the departure draft, which is known only
after service is complete. The predictive accuracies reported for the tree models
(around $11.7$~h MAE) are therefore likely optimistic, and no operational claim
about forecast quality should be made until the models are re-fit without this
feature and any other post-service quantity. A leakage-free re-fit is a
prerequisite for, not a refinement of, the accuracy results.

\emph{Reproducibility.} The exact feature-engineered training table cannot be
regenerated from the released scripts, which limits independent replication of
the predictive results. In addition, the optimizer depends on a
network-metered commercial license, which makes large experiment sweeps slow and
fragile. We mitigated the second issue by also solving with an open-source
solver; the two agree on the optima for the small instances used here, so the
decision results do not hinge on the commercial solver.

\emph{Experimental scale.} Instance counts, training epochs, and seed counts were
kept modest for computational tractability. The synthetic contention effect in
particular rests on a limited number of seeds and should be read as preliminary
evidence of a mechanism rather than a precise estimate of its magnitude.

\emph{Objective scope.} The deployed objective is unweighted. Vessel-priority
weights, which would let the optimizer, and through it the decision-focused loss,
prioritize high-value calls, are deferred to future work. The present study
therefore isolates the cascade-asymmetry channel and does not speak to whether
priority weighting would enlarge or shrink the DFL advantage.

\emph{Single port, single period.} All results come from one port over one
period. The operational regularities that shape both the features and the
contention structure are specific to that terminal, and generalization to ports
with different berth layouts, traffic mixes, or service-time distributions is
untested.

\subsection{Practical implications}\label{sec:disc-practice}

The overall message is deliberately cautious. On this evidence, decision-focused
learning is not a turnkey improvement for berth scheduling. Its advantage is
real where we can observe the mechanism at work, but it is conditional on high
enough contention that prediction errors cascade, on a training budget sufficient
to express the hedge, on temporally coherent instances that sustain that
contention, and on a stable gradient through the scheduling program. It also
rests, as does every accuracy figure here, on leakage-free features. When those
conditions are absent, as in our real-data experiment, DFL matches the
predict-then-optimize baseline at best and can trail it when the gradient
diverges.

The path forward is to address the gradient and the instances rather than to
abandon the approach. Feasibility-aware formulations of DFL are designed for the
setting where predictions enter constraints and offer a more stable training
signal than a plain black-box estimator \citep{mandi2025feasibility}. Perturbed
and implicit differentiation of the decision layer
\citep{berthet2020learning, niepert2021imle} provide lower-variance gradients
that may remove the need for aggressive early stopping and let training move
meaningfully beyond the warm start. Pairing either of these with genuinely
coherent weekly instances, once such data can be assembled, would test the
cascade-asymmetry hypothesis under the conditions where the synthetic results
suggest it should matter most.
