\section{Introduction}\label{sec:intro}

Seaport container terminals allocate arriving vessels to a limited set of
heterogeneous berths, and the quality of that allocation governs how long ships
wait, how evenly quay cranes are used, and how reliably the terminal meets its
service commitments. The berth allocation problem (BAP) formalizes this
decision, and it has been studied extensively over the last two decades
\citep{bierwirth2010survey}. At a port such as San Antonio, Chile, the country's
largest container gateway, the difficulty is compounded by a shared entry
channel that vessels must transit one at a time and by berths that differ in
length, draft, and crane equipment. A planner must decide, for a week of
expected arrivals, which vessel occupies which berth and in what order, so that
later arrivals are not held offshore while an earlier vessel is still working.
The channel couples the berths: a vessel cannot enter or leave while another is
transiting, so a schedule that looks feasible berth by berth can still stall
when the shared passage is contested. Small errors in the assumed timing of one
vessel therefore ripple beyond its own berth.

The load-bearing input to this decision is the expected service time of each
vessel, that is, the time a ship will occupy its berth from mooring to
departure. Every scheduling model consumes these durations, and the schedule it
returns is only as good as the durations it was given. Service time is not
known in advance; it depends on the volume of containers to move, the crane
allocation, the labor shift, and conditions that are realized only once the
vessel is alongside. Because service time is uncertain and consequential, a
growing body of work predicts it from historical operational data and treats
the prediction as a plug-in for downstream planning
\citep{yan2024predicting,chu2025vessel}, while a parallel line of work hedges
against the uncertainty directly through robust and stochastic formulations
\citep{rodrigues2022uncertainty,kolley2023robust}. Both directions share a
common premise: a better estimate of service time yields a better schedule.

The prevailing predictive pipeline is two-stage. A machine-learning model is
first trained to minimize a prediction-error metric such as mean absolute or
root-mean-squared error, and its point predictions are then passed, unchanged,
to the optimizer that builds the schedule. This predict-then-optimize (PtO)
arrangement is convenient because the two stages can be developed and validated
independently, but it optimizes the wrong quantity: forecast accuracy is a
proxy for what the planner actually cares about, which is the cost of the
schedule the forecast induces. Decision-focused learning (DFL) removes the
proxy by training the predictor on the downstream decision cost, or regret,
rather than on prediction error, so that the model learns to be accurate where
accuracy changes the decision and is permitted to be imprecise where it does
not \citep{elmachtoub2022spo,mandi2024decision}. DFL has been applied to
shortest-path, knapsack, and portfolio problems, but to our knowledge it has not
been applied to berth allocation. It has also rarely been studied in the
setting that berth allocation demands, where the prediction enters the
scheduling \emph{constraints} rather than a linear objective cost vector
\citep{hu2023twostage,mandi2025feasibility}. This distinction matters: the
Smart Predict-then-Optimize (SPO+) surrogate of \citet{elmachtoub2022spo}
assumes the prediction is the objective's cost vector and does not apply when
the prediction sits inside a precedence constraint, as a service time does.

The reason to expect DFL to help here is that schedule cost is asymmetric in the
prediction error, and a squared-error loss is blind to that asymmetry. Consider
a berth working a sequence of vessels. If the model under-predicts a vessel's
service time, the optimizer schedules the next vessel to moor too early; when
the vessel in fact overruns, the following one is delayed, and that delay
propagates down the queue as each subsequent vessel inherits the accumulated
slippage. If the model over-predicts, the optimizer leaves the berth idle for a
short interval and the next vessel simply starts on time, with no downstream
consequence. The two errors are of equal magnitude but not of equal cost: one
cascades and the other does not. Mean-squared error penalizes over- and
under-prediction identically, so a PtO predictor has no incentive to err on the
conservative side. A decision-focused loss, computed through the scheduler, sees
the cascade directly and can push the predictor toward the errors that are cheap
to make. We refer to this mechanism as \emph{cascade asymmetry}, and it is the
hypothesis that motivates the present study.

We test this hypothesis empirically, and we report a mixed outcome rather than a
clean win. The contributions of this paper are the following. First, we
formulate a discrete, weekly-horizon BAP as a mixed-integer linear program and
integrate a learned service-time predictor into it, differentiating through the
solver with a black-box scheme
\citep{pogancic2020blackbox,berthet2020learning,tang2024pyepo}. To our
knowledge this is the first decision-focused treatment of berth allocation, and
it operates in the harder regime where the prediction enters the model's
constraints rather than its objective coefficients. Second, we make a
formulation choice that aligns the training loss and the reported regret metric
with the solver's own objective, which yields an exact guarantee of
non-negative regret; we verify this guarantee both analytically and by
enumeration over small instances. Third, we evaluate the approach on a real
dataset of vessel calls at the Port of San Antonio spanning 2019 to 2025 and on
controlled synthetic instances. On the synthetic instances DFL reduces schedule
regret relative to PtO, and the margin widens as berth contention increases,
which is consistent with the cascade-asymmetry mechanism. On the real-data
instances the gain falls within run-to-run variance, and the black-box gradient
is unstable, so that a usable result requires careful learning-rate control and
early stopping. We do not claim that DFL robustly outperforms PtO on real data.
Fourth, we give an honest account of the practical obstacles that a deployable
DFL-for-BAP system would have to overcome, including the gradient instability
noted above, a feature we suspect leaks the target, and limitations in the
geometry of the available instances that blunt the decision signal DFL relies
on.

The remainder of the paper is organized as follows. Section~\ref{sec:related}
reviews the berth allocation literature, the treatment of service-time
uncertainty, and decision-focused learning, and locates our contribution within
them. Section~\ref{sec:problem} describes the operational setting at the Port of
San Antonio and the vessel-call dataset, including its limitations.
Section~\ref{sec:method} presents the discrete BAP model, the service-time
predictors, and the decision-focused training procedure.
Section~\ref{sec:results} reports the computational experiments on synthetic and
real instances. Section~\ref{sec:discussion} interprets the results and the
practical obstacles they expose, and Section~\ref{sec:conclusion} concludes.
