\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[round]{natbib}
\usepackage{hyperref}

% Color palette (shared with docs/meetings/dfl_explainer.tex)
\definecolor{predBlue}{HTML}{2C6FA8}
\definecolor{optOrange}{HTML}{D97706}
\definecolor{regretRed}{HTML}{B23A48}

\newcommand{\I}{\mathcal{I}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\Bi}{\mathcal{B}_i}
\newcommand{\bigM}{M}

\title{A Week-long Discrete Berth Allocation MILP with No-wait Priority
``Services'' and Vessel--Berth Compatibility, for Decision-Focused Learning}
\author{portsDFL --- Port of San Antonio}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We formulate the operational problem of planning one week of berth assignments
at the Port of San Antonio as a discrete, dynamic Berth Allocation Problem
(BAP). The model assigns each vessel to a compatible berth and a start time,
minimizing total weighted completion time. A subset of \emph{service} vessels
(the port's term for committed/priority calls) carry a \emph{hard} no-wait
window: they must be berthed essentially on arrival. An optional shared
\emph{navigation channel}---a single waterway every vessel transits to enter and
to exit, with no two transits overlapping---can be switched on, in which case the
model minimizes weighted \emph{departure}. The service time
$\tau$ that drives the schedule is predicted by an upstream model and the
predictor is trained \emph{decision-focused}, differentiating end-to-end
through the MILP, with a predict-then-optimize baseline as the natural
ablation. This note states the formulation, its hard/soft window modes, the
single-week solve scope, and how it sits in the published BAP literature.
\end{abstract}

\section{Problem setting}
The Port of San Antonio operates several terminals/berths that are
\emph{heterogeneous}: each can physically handle only certain vessel types
(e.g.\ liquid bulk only at the liquid-bulk berth). Over a planning week, vessels
arrive at known times and must each be assigned to a compatible berth and a
start time. Two operational realities drive the formulation:
\begin{itemize}[nosep]
  \item \textbf{Heterogeneous berths.} A vessel may use only the berths
  compatible with its type; the compatibility map is populated data-drivenly
  from the historical $(\text{terminal}\times\text{vessel-type})$ table.
  \item \textbf{Priority ``services''.} A subset of vessels are committed calls
  that must not wait: they are berthed at (or within a small slack of) their
  arrival time. This is a hard reservation, not a soft preference.
\end{itemize}
The service time $\tau_i$ (berth occupation) is not known in advance; it is
predicted from vessel/observed features. The optimizer consumes the prediction
$\hat\tau$, and the predictor is trained so its predictions yield good
\emph{decisions} (\S\ref{sec:dfl}).

\section{Sets, parameters, and variables}
\begin{table}[h]
\centering
\small
\begin{tabular}{@{}ll@{}}
\toprule
Symbol & Meaning \\
\midrule
$\I=\{1,\dots,N\}$ & vessels in the planning week \\
$\B=\{1,\dots,B\}$ & berths \\
$\Bi\subseteq\B$ & berths \emph{compatible} with vessel $i$ \\
$\R\subseteq\I$ & ``service'' (priority, no-wait) vessels \\
$a_i\ge 0$ & arrival time of vessel $i$ (hours from week start) \\
$w_i\ge 0$ & priority weight of vessel $i$ \\
$\tau_i>0$ & service (berth-occupation) time; \emph{predicted} as $\hat\tau_i$ \\
$l_r$ & latest start for service $r\in\R$; $l_r=a_r+\text{slack}$ \\
$c\ge 0$ & navigation-channel transit time (enter or exit); \emph{optional} \\
$\bigM$ & big-$M$ constant ($\ge$ horizon $+$ max berth workload; $+2Nc$ with channel) \\
\midrule
$x_{ib}\in\{0,1\}$ & $1$ if vessel $i$ uses berth $b$ \quad(only for $b\in\Bi$) \\
$s_i\ge a_i$ & start time of vessel $i$ \\
$z_{ijb}\in\{0,1\}$ & $1$ if $i$ precedes $j$ at berth $b$ \\
$\delta_r\ge 0$ & (soft mode) tardiness of service $r$ past its window \\
$e^{\mathrm{in}}_i\ge a_i$ & (channel) inbound-transit start of vessel $i$ \\
$e^{\mathrm{out}}_i\ge 0$ & (channel) outbound-transit start of vessel $i$ \\
$y_{ef}\in\{0,1\}$ & (channel) $1$ if transit event $e$ precedes event $f$ \\
\bottomrule
\end{tabular}
\caption{Notation.}
\end{table}

\section{Formulation}
The discrete BAP, time-precedence variant
\citep{cordeau2005berth, alzaabi2016berth}:
\begin{align}
\min\quad & \sum_{i\in\I} w_i\,(s_i + \tau_i)
   \;\;\big[\,+\;p\!\!\sum_{r\in\R}\delta_r\ \text{(soft mode)}\,\big]
   \label{eq:obj}\\
\text{s.t.}\quad
& \sum_{b\in\Bi} x_{ib} = 1 && \forall i\in\I \label{eq:assign}\\
& s_i \ge a_i && \forall i\in\I \label{eq:arrival}\\
& s_r \le l_r && \forall r\in\R \ \text{(hard mode)} \label{eq:window}\\
& \delta_r \ge s_r - l_r && \forall r\in\R \ \text{(soft mode)} \label{eq:soft}\\
& z_{ijb} + z_{jib} \le 1 && \forall b,\ i<j:\ i,j\in\Bi\cap\B_j \label{eq:one}\\
& z_{ijb} + z_{jib} \ge x_{ib} + x_{jb} - 1 && \forall b,\ i<j \label{eq:order}\\
& s_j \ge s_i + \tau_i - \bigM\,(1 - z_{ijb}) && \forall (i,j,b),\ i\ne j \label{eq:prec}
\end{align}
\eqref{eq:obj} is total weighted completion time. \eqref{eq:assign} assigns each
vessel to exactly one \emph{compatible} berth---incompatible $x_{ib}$ are never
created (the data-driven analogue of the forbidden-assignment cut
$X_{ij}\le d_{ij}$ of \citealp{alzaabi2016berth}). \eqref{eq:arrival} is the
dynamic arrival release. \eqref{eq:window}/\eqref{eq:soft} encode the no-wait
service window (see \S\ref{sec:windows}). \eqref{eq:one}--\eqref{eq:order} force
an ordering between any two vessels sharing a berth, and \eqref{eq:prec} is the
big-$M$ precedence that prevents overlap. Precedence triples and $z$ variables
are built only for pairs that can share a berth ($b\in\Bi\cap\B_j$), which keeps
the model compact under tight compatibility. When the optional navigation
channel is modeled (\S\ref{sec:channel}), objective \eqref{eq:obj} is replaced by
weighted \emph{departure} $\sum_{i\in\I} w_i\,(e^{\mathrm{out}}_i + c)$, since the
vessel only clears the port after its outbound transit completes.

\section{Hard vs.\ soft service windows}\label{sec:windows}
\textbf{Hard mode} (deterministic weekly planner) imposes \eqref{eq:window}:
a service vessel literally cannot start after $l_r$. With $\text{slack}\to 0$
this means ``berth on arrival.'' An over-subscribed week (more concurrent
no-wait services than compatible berths) is then \emph{infeasible}---a genuine
planning signal, surfaced to the planner rather than silently absorbed.

\textbf{Soft mode} (decision-focused training) drops \eqref{eq:window} and adds
the tardiness variable and penalty $\eqref{eq:soft}$, $+\,p\sum_r\delta_r$ in
\eqref{eq:obj}. This keeps every solve \emph{feasible} under any predicted
$\hat\tau$---essential because the DFL trainer re-solves the MILP for arbitrary
(early-training, possibly poor) predictions, and an infeasible solve would abort
training and undefine the regret.

\section{Shared navigation channel}\label{sec:channel}
The Port of San Antonio has a \emph{single} navigation channel: every vessel
transits it once to \textbf{enter} (before berthing) and once to \textbf{exit}
(after service), and no two transits---of any vessels, in either
direction---may occupy it simultaneously. This is an optional extension: with
$c=\text{None}$ the model reduces exactly to the berth-only formulation above.

Each vessel gets an inbound-transit start $e^{\mathrm{in}}_i\ge a_i$ and an
outbound-transit start $e^{\mathrm{out}}_i$; transit $k\in\{\mathrm{in},
\mathrm{out}\}$ occupies the channel over $[t^k_i,\,t^k_i+c]$ with
$t^{\mathrm{in}}_i=e^{\mathrm{in}}_i$ and $t^{\mathrm{out}}_i=e^{\mathrm{out}}_i$.
Let $\mathcal{E}=\{(i,\mathrm{in}),(i,\mathrm{out}):i\in\I\}$ be the $2N$ transit
events. The channel adds:
\begin{align}
& s_i \ge e^{\mathrm{in}}_i + c && \forall i\in\I \label{eq:chin}\\
& e^{\mathrm{out}}_i \ge s_i + \tau_i && \forall i\in\I \label{eq:chout}\\
& t_f \ge t_e + c - \bigM\,(1-y_{ef}) && \forall\, e,f\in\mathcal{E},\ e<f,\ v(e)\ne v(f) \label{eq:chfwd}\\
& t_e \ge t_f + c - \bigM\,y_{ef} && \forall\, e,f\in\mathcal{E},\ e<f,\ v(e)\ne v(f) \label{eq:chbwd}
\end{align}
\eqref{eq:chin} mooring may begin only after the inbound transit finishes;
\eqref{eq:chout} the outbound transit may begin only after service completes.
\eqref{eq:chfwd}--\eqref{eq:chbwd} are the channel mutual-exclusion disjunction:
for every pair of transit events belonging to \emph{different} vessels ($v(e)$ is
$e$'s vessel), one transit must finish before the other starts. The same-vessel
pair $(i,\mathrm{in}),(i,\mathrm{out})$ is omitted---it is already ordered by
\eqref{eq:chin}--\eqref{eq:chout}---giving $\binom{2N}{2}-N$ order binaries
$y_{ef}$. The disjunction big-$M$ uses $\bigM+2Nc$, large enough for the channel
to serialize all $2N$ transits.

\textbf{Objective.} With the channel, the model minimizes weighted
\emph{departure} $\sum_{i\in\I} w_i\,(e^{\mathrm{out}}_i + c)$: the vessel leaves
the port only when its outbound transit clears the channel.

\textbf{No-wait window on entry.} When the channel is modeled, the hard/soft
service window \eqref{eq:window}/\eqref{eq:soft} is imposed on the
\emph{entry} time ($e^{\mathrm{in}}_r\le l_r$ hard, $\delta_r\ge
e^{\mathrm{in}}_r-l_r$ soft) rather than on $s_r$: ``no waiting'' for a priority
service means it enters the channel on arrival, the mandatory transit time $c$
notwithstanding.

\textbf{DFL scope.} The channel defaults off for the synthetic generators and
the DFL trainer, so the PyEPO contract (cost$\leftrightarrow$decision of length
$N$) is unchanged; the deterministic weekly planner is its consumer. Extending
the regret re-derivation to model the channel is a documented follow-up, not in
scope here.

\section{Single-week solve scope}
The MILP solves exactly one 7-day instance ($N$ fixed, horizon $168$\,h).
Selecting the week's vessels is a \emph{pre-solve slicing step}; there is no
multi-week, rolling-horizon, or cross-week carry-over inside the optimizer.
Vessels arriving outside the chosen window are filtered out before the model is
built.

\section{Decision-focused learning integration}\label{sec:dfl}
$\hat\tau$ enters both the objective \eqref{eq:obj} and the precedence
constraints \eqref{eq:prec}. Because the predicted quantity appears in the
constraints (not only as a linear objective cost vector), SPO+
\citep{elmachtoub2022spo} does not apply directly; we differentiate through the
solver with PyEPO blackbox differentiation \citep{pogancic2020blackbox}, which
trains the predictor to minimize downstream regret. A predict-then-optimize
(MSE) predictor is the controlled baseline. Regret is measured against the
full-information schedule (the MILP solved under true $\tau$), and is
$\ge 0$ by construction.

\input{related_formulations}

\section{Implementation pointers}
The MILP lives in \texttt{optimizers/src/bap\_optim/discrete\_bap.py}
(\texttt{DiscreteBAP}, with the \texttt{hard\_windows} flag; \texttt{extract\_channel}
reads back $e^{\mathrm{in}}/e^{\mathrm{out}}$). The instance
descriptor \texttt{BAPInstance} (with optional \texttt{latest\_start},
\texttt{berth\_compat}, \texttt{service}, \texttt{channel\_time}) is in \texttt{bap\_optim/instance.py}; the
berth catalog and compatibility in \texttt{bap\_optim/berths.py}; the pre-solve
weekly slicing in \texttt{bap\_optim/weekly\_instance.py}; and the deterministic
planner in \texttt{prediction\_models/scripts/plan\_week.py}. Open modeling decisions
and assumptions are tracked in \texttt{prediction\_models/docs/formulation/decisions\_and\_questions.md}.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
